Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.

### 2023-10-19 5_artigos/the-international-journal-of-biostatistics/draft/TABLES/8-table8.tex
adiciona artigo ijb

\documentclass[AMA,STIX1COL]{WileyNJD-v2}

\articletype{RESEARCH ARTICLE}%

\received{Day April 2016}
\revised{6 June 2016}
\accepted{6 June 2016}

\raggedbottom

\begin{document}

\title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.}

\author[1]{Lineu Alberto Cavazani de Freitas}

\author[2]{Ligia de Oliveira Carlos}

\author[2]{Antônio Carlos Ligocki Campos}

\author[3]{Wagner Hugo Bonat}

\authormark{FREITAS \textsc{et al}}

\address[1]{\orgdiv{Department of Informatics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[2]{\orgdiv{Department of Surgical Clinic}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[3]{\orgdiv{Department of Statistics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\corres{Lineu Alberto Cavazani de Freitas, Department of Informatics, Paraná Federal University, R. Evaristo F. Ferreira da Costa, 383-391, Jardim das Américas, Curitiba, Paraná, 82590-300, Brazil. \email{lineuacf@gmail.com}}

\abstract[Summary]{

Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

}

\keywords{McGLM, Hypothesis tests, Wald test, ANOVA, MANOVA, Multiple comparisons}

\jnlcitation{\cname{%
\author{Freitas LAC}, 
\author{Carlos LO},
\author{Campos ACL} and 
\author{Bonat WH}} (\cyear{2022}), 
\ctitle{Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models}, \cjournal{Statistics in Medicine}}

\maketitle

%\footnotetext{\textbf{Abbreviations:} ANA, anti-nuclear antibodies; APC, antigen-presenting cells; IRF, interferon regulatory factor}

%-----------------------------------------------------------------------

\section{Introduction}\label{sec1}

% Estudos clínicos

Clinical trials represent one of the most common procedures used to assess whether a new treatment is safe, effective or even better than a standard treatment. These are procedures widely used in medical research with the aim of evaluating interventions on a group of individuals. In general, the data generated by clinical trials are not trivial and, for analysis, there is a need for statistical methods suitable to deal with all the specifics of this type of study.

According to Meinert and Tonascia \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

In this type of study participants are randomly assigned to a treatment or control group (or to multiple treatment groups) \cite{hannan2008randomized} and the objective is to assess whether there is a significant difference between the groups.

% Problemas na área médica que envolvem múltiplas respostas

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: Kangovi et al \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; Song and Baicker \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; Thyregod et al \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); Schmitz et al \cite{schmitz2019effect} evaluated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; Bonat et al \cite{plastica} proposed a model to deal with multiple outcomes in repeated measures studies motivated by a problem that aimed to compare different aesthetic eyelid surgery techniques; Petterle et al \cite{petterle2021multivariate} proposed a multivariate regression model to handle multiple continuous bounded outcomes motivated by a body fat percentage data. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

% Introdução ao Dataset

Similarly to the aforementioned articles, we are interested in the analysis of multiple outcomes in the context of longitudinal data analysis. The study we shall describe was conducted with the objective of evaluating the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. This is a clinical trial with two response variables: a score that characterizes compulsion and the number of symptoms presented that characterize addiction. Both responses are limited counts and for analysis purposes have been transformed to unit range.

In this study, a set of subjects was divided into two groups: one of them received a placebo and the other received the treatment. In addition, individuals were evaluated at 3 different time points: the first evaluation was performed approximately 10 days before surgery; follow-up assessments were performed approximately three months and one year postoperatively.

% McGLM

It is important to note that this is a multivariate problem (contains two response variables), with non-Gaussian responses (both taking values in the unit range) and that the observations that make up the data set cannot be considered independent, considering that measurements taken on the same individual at different time points tend to be correlated. Therefore, it is a problem that standard modeling techniques would be difficult to apply. It is necessary to use some methodology that supports the design of the clinical trial adequately.

In this article, we adopt the multivariate covariance generalized linear models (McGLM) framework \cite{Bonat16}, which provides an environment for modeling multiple non-Gaussian responses simultaneously with flexible and interpretable modeling of the covariance structure. The within outcomes covariance matrix is specified for each marginal outcome using a linear combination of known matrices, while the joint covariance matrix is specified using the generalized Kronecker product \cite{martinez13,Bonat16}.

% Importância de testes de hipóteses em regressão

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. Engle \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

% ANOVA e MANOVA

In the case of standard linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by Fisher and Mackenzie \cite{anova_fisher}. According to St, Wold et al \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and it is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. Among the multivariate hypothesis tests already discussed in the literature for MANOVAs, we have the Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

% Testes de comparações múltiplas

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences appear. For example, suppose there is a three-level categorical variable $X$ in the model that assumes the levels: A, B, or C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used, some of the tests discussed in the literature are: the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. Hsu \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. Bretz, Hothorn, and Westfall \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

% Objetivo do trabalho

However, for multiple response regression models, there are few alternatives for hypothesis testing. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is the development of hypothesis tests for McGLMs. We have the following specific objectives: to propose the use of the Wald test to carry out tests of general hypotheses on regression and dispersion parameters of McGLMs, also enabling the generation of tables of analysis of variance, multivariate analysis of variance and multiple comparisons tests for regression models with multiple Gaussian and non-Gaussian responses. Another objective is to evaluate the properties of the proposed tests based on simulation studies and to motivate the potential application of the discussed methodologies based on the analysis of the aforementioned dataset. In the case study, standard modeling techniques would be difficult to apply, however it is a problem of possible analysis via McGLM and hypothesis tests can be used to evaluate the effect of the interaction between time and use of the probiotic on addiction and binge eating disorder.

% Organização do trabalho (seções)

We present the dataset in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations. The data sets and R codes are available online and can be assessed in \href{https://github.com/lineu96/paper-htmcglm}{https://github.com/lineu96/paper-htmcglm}.

%-----------------------------------------------------------------------

\section{Data set}\label{sec2}

% Contexto

This is a randomized, double-blind, placebo-controlled clinical trial conducted with patients undergoing Roux-en-Y Gastric Bypass (RYGB) from April 2018 to December 2019. The study was approved by the Research Ethics Committee of the Pontifical Catholic University of Paraná (PUCPR) (nº 4.252.808 ) and registered by the Brazilian Registry of Clinical Trials - REBEC (nºRBR-4x3gqp). The research was explained to each participant before their participation and, from those who agreed, written informed consent was obtained. The division of groups (placebo or probiotic) was done randomly. The inclusion criteria for subjects in the study were: adults (18-59 years) who would undergo RYGB, with a body mass index (BMI) $\geq$ 35 kg/m2 and who did not use antibiotics before surgery.

Patients who underwent other surgical techniques or reoperation, had post-surgical complications, received antibiotic therapy concomitantly with the use of probiotic/placebo or did not use probiotic/placebo tablets for more than nine consecutive days were excluded from the study. Both groups received the same dietary guidelines after surgery, were followed up by the same surgical team (doctor, nutritionist and psychologist) and had the same number of pre-scheduled consultations before and after surgery, following the protocol established by the institution where the study was carried out.

On the seventh postoperative day, participants were instructed to ingest two chewable tablets/day of placebo or Flora Vantage probiotic tablet, 5 billion of Lactobacillus acidophilus NCFM \textregistered Strain and 5 billion of Bifidobacterium lactis Bi-07 \textregistered) from Bariatric Advantage (Aliso Viejo, CA, USA) for 90 days. Subjects were evaluated at 3 time points. The first evaluation (T0) was performed approximately 10 days before surgery. Follow-up assessments were performed approximately three months (T1) and one year postoperatively (T2). At these moments, clinical and anthropometric assessments were performed, as well as self-administered questionnaires were given to the participants at each meeting.

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by Gormally et al \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by Gearhardt et al \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

% Desenho experimental e coleta de dados

The final sample consists of 71 individuals: 33 belong to the placebo group and 38 to the treatment group. If all these individuals were evaluated at the 3 time points defined in the study, the dataset would have 213 observations. However, throughout the study, several individuals did not attend the consultations, which causes missing data in the dataset. After processing the data and excluding missing observations, 184 observations remained.

% Conjunto de dados

For analysis purposes, the score that characterizes compulsion and the number of symptoms presented that characterize addiction were transformed to the unitary scale, considering that these are counts variables restricted to a range. The purpose of the analysis is to assess the effect of moment and group on addiction and compulsion metrics. The dataset contains the following variables:

\begin{itemize}
  \item id: individual identifier variable.
  \item moment: moment identifier variable (T0, T1, T2).
  \item group: group identifier variable (placebo, probiotic).
  \item YFAS: proportion of symptoms that characterize addiction.
  \item BES: BES score transformed to unit scale.
\end{itemize}

% Análise exploratória

The graphical analysis presented in \autoref{fig1} shows, in (a) and (d) that both variables of interest present considerable asymmetry to the right. The boxplots of metrics as a function of group presented in (b) and (e) show little difference between the placebo and probiotic groups for both responses. The boxplots of the metrics as a function of the evaluation moments, presented in (c) and (f), show that for both metrics the values were higher at T0, with a considerable reduction at T1. When comparing T1 and T2, for YFAS it seems that there is a slight increase in the last evaluation; for BES, T1 and T2 do not seem to differ.

\begin{figure}[h]
\centerline{\includegraphics[width=342pt,height=9pc,draft]{empty}}
\caption{Exploratory graphic analysis: (a) YFAS histogram, (b) YFAS boxplots as a function of group, (c) YFAS boxplots as a function of moment, (d) BES histogram, (b) BES boxplots as a function of group, (c) BES boxplots as a function of moment. The asterisk in the boxplots indicates the mean.\label{fig1}}
\end{figure}

Still in an exploratory way, we can evaluate the behavior of addiction and compulsion metrics through the evaluation of descriptive measures by moment and by group, presented in \autoref{tab:tab1}. It is possible to verify the reduction of individuals over time, something common in prospective studies. Regarding the measurements, it is noted that both groups (placebo or probiotic) have higher means at T0 than at other times. Therefore, there is a clear reduction in metrics when compared to the preoperative period. When comparing the postoperative moments (T1 and T2) we verified that the YFAS measurements for the placebo and probiotic groups showed, on average, an increase in the metrics at the last evaluation moment. The same could be seen for BES in the placebo group. On the other hand, the BES measures in the probiotic group showed a decrease.

%-----------------------------------------------------------------------

\section{Multivariate Covariance Generalized Linear Models}\label{sec3}

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ be a matrix of response variables and $\boldsymbol{ M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ be a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$, is denoted by $\Sigma_r$ and has dimension $N \times N$. In addition, consider a correlation matrix $\Sigma_b$, of order $R \times R$, which describes the correlation between the response variables.

The McGLMs \citep{Bonat16} are defined by:

$$
      \begin{aligned}
        \mathrm{E}(\boldsymbol{Y}) &=
          \boldsymbol{M} =
            \{g_1^{-1}(\boldsymbol{X}_1 \boldsymbol{\beta}_1),
            \ldots,
            g_R^{-1}(\boldsymbol{X}_R \boldsymbol{\beta}_R)\}
          \\
        \mathrm{Var}(\boldsymbol{Y}) &=
          \boldsymbol{C} =
            \boldsymbol{\Sigma}_R \overset{G} \otimes
            \boldsymbol{\Sigma}_b,
      \end{aligned}
$$

\noindent where the $g_r()$ are standard link functions; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

For continuous, binary, binomial, proportions or indices, the variance and covariance matrix $\boldsymbol{\Sigma}_r$ is given by:

$$
\Sigma_r =
\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right)^{1/2}(\boldsymbol{\Omega}\left(\boldsymbol{\tau}_r\right))\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right)^{1/2}.
$$

In the case of response variables that are counts, the variance and covariance matrix for each response variable is given by:

$$
\Sigma_r = diag(\boldsymbol{\mu}_r)+ \mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right)^{1/2}(\boldsymbol{\Omega}\left(\boldsymbol{\tau}_r\right))\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right)^{1/2},
$$

\noindent where $\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right) = diag(\vartheta(\boldsymbol{\mu}_r; p_r))$ denotes a diagonal matrix in which the entries are given by the variance function $\vartheta(\cdot; p_r)$ applied to the elements of the vector $\boldsymbol{\mu}_r$. Different choices of variance functions $\vartheta(\cdot; p_r)$ imply different assumptions about the distribution of the response variable. We will mention 3 options of variance functions: power variance function, Poisson–Tweedie dispersion function and binomial variance function.

The power variance function characterizes the Tweedie family of distributions, it is given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_r}_r$, in which some distributions stand out: Normal ($p $ = 0), Poisson ($p$ = 1), gamma ($p$ = 2) and Inverse Gaussian ($p$ = 3) \cite{Jorgensen87, Jorgensen97}.

The Poisson–Tweedie dispersion function \cite{Jorgensen15} is an option for responses that characterize counts. The dispersion function given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. Thus, we have a rich class of models to deal with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

Finally, the binomial variance function, given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_{r1}}_r(1 - \mu_r)^{p_{r2}}$ is indicated when the response variable is binary, bounded to an interval, or when there is a number of successes in a number of trials.

It is possible to notice that the power parameter $p$ appears in all the variance/dispersion functions discussed. This parameter is especially important because it is an index that distinguishes among different probability distributions that are important in the modeling context and, for this reason, can be used as a tool for automatic selection of the probability distribution that best fits to the data.

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of Anderson et al \cite{Anderson73} and Pourahmadi \cite{Pourahmadi00}, Bonat and Jørgensen \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

$$
h\left \{ \boldsymbol{\Omega}(\boldsymbol{\tau}_r) \right \} = \tau_{r0}Z_0 + \ldots + \tau_{rD}Z_D,
$$

\noindent where $h()$ is the covariance link function, $Z_{rd}$ with $d$ = 0,$\ldots$, D are known matrices representing the covariance structure present in each response variable $r$ and $\boldsymbol{\tau_r}$ = $(\tau_{r0}, \ldots, \tau_{rD})$ is a vector $(D + 1) \times 1$ of dispersion parameters.

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by Pinheiro and Bates \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most well-known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by Demidenko \cite{Demidenko13}.

In this way, the McGLMs configure a general framework for analysis via regression models for Gaussian and non-Gaussian data with multiple responses in which no assumptions are made regarding the independence of the observations. The class is defined by 3 functions (link, variance and covariance) in addition to a linear predictor and a matrix linear predictor for each response under analysis.

\subsection{Estimation and inference}

McGLMs are fitted based on the estimating function approach described in detail by Bonat and Jørgensen \cite{Bonat16} and Jørgensen and Knudsen \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

McGLMs' second-moment assumptions allow us to spplit the parameter vector into two sets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^ {\top}$. So $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a $K \times 1$ vector of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top , \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a $Q \times 1$ vector of dispersion parameters. Also, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector of order $NR \times 1$ from the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^ \top)^\top$ denotes the stacked vector of order $NR \times 1$ of the expected value matrix $\boldsymbol{M}_{N \times R}$.

To estimate the regression parameters, the following quasi-score function \cite{Liang86} is used, 
$$
\begin{aligned}
  \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
  \boldsymbol{\lambda}) = \boldsymbol{D}^\top
  \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
\end{aligned}
$$

\noindent where $\boldsymbol{D} = \nabla_{\boldsymbol{\beta}} \mathcal{M}$ is a $NR \times K$ matrix, and $\nabla_{\boldsymbol{\beta}}$ denotes the gradient operator. Using the quasi-score function the $K \times K$ sensitivity matrix of $\psi_{\boldsymbol{\beta}}$ is given by

$$
\begin{aligned}
S_{\boldsymbol{\beta}} = E(\nabla_{\boldsymbol{\beta} \psi \boldsymbol{\beta}}) = -\boldsymbol{D}^{\top} \boldsymbol{C}^{-1} \boldsymbol{D},
\end{aligned}
$$

\noindent the $K \times K$ variability matrix of $\psi_{\boldsymbol{\beta}}$ is written as

$$
\begin{aligned}
V_{\boldsymbol{\beta}} = VAR(\psi \boldsymbol{\beta}) = \boldsymbol{D}^{\top} \boldsymbol{C}^{-1} \boldsymbol{D}.
\end{aligned}
$$

For the dispersion parameters, the Pearson estimating function is used, defined in the form

$$
  \begin{aligned}
    \psi_{\boldsymbol{\lambda}_i}(\boldsymbol{\beta},
    \boldsymbol{\lambda}) =
    \mathrm{tr}(W_{\boldsymbol{\lambda}i}
    (\boldsymbol{r}^\top\boldsymbol{r} -
    \boldsymbol{C})),  i = 1,.., Q, 
  \end{aligned}
$$

\noindent where $W_{\boldsymbol{\lambda}i} = -\frac{\partial \boldsymbol{C}^{-1}}{\partial \boldsymbol{\lambda}_i}$ and $\boldsymbol{ r} = (\mathcal{Y} - \mathcal{M})$. The entry $(i,j)$ of the sensitivity matrix $Q \times Q$ of $\psi_{\boldsymbol{\lambda}}$ is given by

$$
  \begin{aligned}
    S_{\boldsymbol{\lambda_{ij}}} = E \left (\frac{\partial }{\partial \boldsymbol{\lambda_{i}}} \psi \boldsymbol{\lambda_{j}}\right) = -tr(W_{\boldsymbol{\lambda_{i}}} CW_{\boldsymbol{\lambda_{J}}} C).
  \end{aligned}
$$

\noindent The entry $(i,j)$ of the variability matrix $Q \times Q$ of $\psi_{\boldsymbol{\lambda}}$ is defined by

$$
  \begin{aligned}
V_{\boldsymbol{\lambda_{ij}}} = Cov\left ( \psi_{\boldsymbol{\lambda_{i}}}, \psi_{\boldsymbol{\lambda_{j}}} \right) = 2tr(W_{\boldsymbol{\lambda_{i}}} CW_{\boldsymbol{\lambda_{J}}} C) + \sum_{l=1}^{NR} k_{l}^{(4)} (W_{\boldsymbol{\lambda_{i}}})_{ll} (W_{\boldsymbol{\lambda_{j}}})_{ll},
  \end{aligned}
$$

\noindent where $k_{l}^{(4)}$ denotes the fourth cumulant of $\mathcal{Y}_{l}$. In the McGLM estimation process, empirical versions are used.

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, Bonat and Jørgensen \cite{Bonat16} obtained the sensitivity and cross-variability matrices, denoted by $S_{\boldsymbol{ \lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in Bonat and Jørgensen \cite{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

$$
  \begin{aligned}
    S_{\boldsymbol{\theta}} = \begin{bmatrix}
      S_{\boldsymbol{\beta}} & S_{\boldsymbol{\beta\lambda}} \\ 
      S_{\boldsymbol{\lambda\beta}} & S_{\boldsymbol{\lambda}} 
      \end{bmatrix} \text{e } V_{\boldsymbol{\theta}} = \begin{bmatrix}
      V_{\boldsymbol{\beta}} & V^{\top}_{\boldsymbol{\lambda\beta}} \\ 
      V_{\boldsymbol{\lambda\beta}} & V_{\boldsymbol{\lambda}} 
    \end{bmatrix}.
  \end{aligned}
$$

Let $\boldsymbol{\hat{\theta}} = (\boldsymbol{\hat{\beta}^{\top}}, \boldsymbol{\hat{\lambda}^{\top}})^{\top }$ be the estimator based on estimating functions of $\boldsymbol{\theta}$. Then the asymptotic distribution of $\boldsymbol{\hat{\theta}}$ is

$$
  \begin{aligned}
    \boldsymbol{\hat{\theta}} \sim N(\boldsymbol{\theta}, J_{\boldsymbol{\theta}}^{-1}),
  \end{aligned}
$$

\noindent where $J_{\boldsymbol{\theta}}^{-1}$ is the inverse of the Godambe information matrix, given by $J_{\boldsymbol{\theta}}^{-1} = S_{ \boldsymbol{\theta}}^{-1} V_{\boldsymbol{\theta}} S_{\boldsymbol{\theta}}^{-\top}$, where $S_{\boldsymbol{\theta}} ^{-\top} = (S_{\boldsymbol{\theta}}^{-1})^{\top}.$

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$, Jørgensen and Knudsen \cite{jorg04} proposed the modified Chaser algorithm, which is defined as

$$
\begin{aligned}
\begin{matrix}
\boldsymbol{\beta}^{(i+1)} = \boldsymbol{\beta}^{(i)}- S_{\boldsymbol{\beta}}^{-1} \psi \boldsymbol{\beta} (\boldsymbol{\beta}^{(i)}, \boldsymbol{\lambda}^{(i)}), \\ 
\boldsymbol{\lambda}^{(i+1)} = \boldsymbol{\lambda}^{(i)} - \alpha S_{\boldsymbol{\lambda}}^{-1} \psi \boldsymbol{\lambda} (\boldsymbol{\beta}^{(i+1)}, \boldsymbol{\lambda}^{(i)}).
\end{matrix}
\end{aligned}
$$

%-----------------------------------------------------------------------

\section{Wald Test for McGLMs}\label{sec4}

This section is dedicated to the presentation of our proposal: the use of the Wald test to evaluate McGLMs parameters. It is worth remembering that in McGLMs there are regression, dispersion, power and correlation parameters. Each set of parameters has a relevant practical interpretation in such a way that through the regression parameters it is possible to identify the relevant explanatory variables, through the dispersion parameters it is possible to evaluate the impact of the correlation between units of the data set, through the power parameters it is possible to identify which probability distribution best fits the data according to the variance function, and through the correlation parameters it is possible to assess the association between responses. In this article, we are interested in methods for evaluation of regression and dispersion parameters.

% Hipóteses e estatística de teste

Let $\boldsymbol{\theta^{*}}$ be a $h \times 1$ vector of parameters with the exception of correlation parameters, i.e. $\boldsymbol{\theta^{*}}$ refers only to regression, dispersion or power parameters. 
The parameter estimates of $\boldsymbol{\theta^{*}}$ are given by $\boldsymbol{\hat\theta^{*}}$. Similarly, let $J^{\boldsymbol{*}-1}$ be the $h \times h$ inverse of the Godambe information matrix excluding the correlation parameters. Let $\boldsymbol{L}$ be a $s \times h$ matrix specifying the hypotheses to be tested and $\boldsymbol{c}$ be a $s \times 1$ vector with values under the null hypothesis, where $s$ denotes the number of constraints; the hypotheses to be tested can be written as:

\begin{equation}
\label{eq:hipoteses_wald}
H_0: \boldsymbol{L}\boldsymbol{\theta^{*}} = \boldsymbol{c} \ vs \ H_1: \boldsymbol{L}\boldsymbol{\theta^{*}} \neq \boldsymbol{c}. 
\end{equation}

\noindent In this way, the generalization of the Wald test statistic to verify the validity of a hypothesis about parameters of a McGLM is given by:

$$
W = (\boldsymbol{L\hat\theta^{*}} - \boldsymbol{c})^T \ (\boldsymbol{L \ J^{\boldsymbol{*}-1} \ L^T})^{-1} \ (\boldsymbol{L\hat\theta^{*}} - \boldsymbol{c}),
$$

\noindent where $W \sim \chi^2_s$, that is, regardless of the number of parameters in the hypotheses, the test statistic $W$ is a single value that asymptotically follows the $\chi^2$ distribution with degrees of freedom given by the number of constraints, that is, the number of rows in the matrix $\boldsymbol{L}$, denoted by $s$.

In general, each column of the matrix $\boldsymbol{L}$ corresponds to one of the $h$ parameters of $\boldsymbol{\theta^{*}}$ and each row to a constraint. Its construction basically consists of filling the matrix with 0, 1 and eventually -1 in such a way that the product $\boldsymbol{L}\boldsymbol{\theta^{*}}$ correctly represents the hypotheses of interest. The correct specification of $\boldsymbol{L}$ allows us testing any parameter individually or even formulating hypotheses for several parameters.

In a practical context, after obtaining the estimates of the model parameters, we may be interested in three types of hypotheses: (i) interest in assessing whether there is evidence that allows us to state that only a single parameter is equal to a postulated value; (ii) interest in assessing whether there is evidence to state that a set of parameters is equal to a postulated vector of values; (iii) interest in knowing if the difference between the effects of two variables is equal to 0, that is, if the effect of the variables on the response is the same.

For the purposes of illustrating the types of hypotheses mentioned, consider the situation in which you want to investigate whether a numeric variable $X_1$ has an effect on two response variables, denoted by $Y_1$ and $Y_2$. For this task, consider that a sample with $N$ observations was collected and for each observation the values of $X_1$, $Y_1$ and $Y_2$ were recorded. Based on the collected data, a bivariate McGLM was fitted, with a linear predictor given by:

\begin{equation}
\label{eq:pred_ex}
g_r(\mu_r) = \beta_{r0} + \beta_{r1} X_1, r=1,2,
\end{equation}

\noindent where the index $r$ denotes the response variable, r = 1,2; $\beta_{r0}$ represents the intercept; $\beta_{r1}$ a regression parameter associated with a variable $X_1$. We assume that each response has only one dispersion parameter $\tau_{r0}$ and that the power parameters have been fixed. Therefore, it is a problem in which there are two response variables and only one explanatory variable. We assume that the observations under study are independent, so $Z_0 = I$.

In this scenario, questions of interest could be: is there an effect of the variable $X_1$ on only one of the responses? Is it possible that the variable $X_1$ has an effect on both responses at the same time? Is it possible that the effect of the variable is the same for both responses? All these questions can be answered by testing hypotheses on the model parameters and specified using \autoref{eq:hipoteses_wald}. In the next subsections, we present the elements necessary to conduct each test.

\subsection{Example 1: hypothesis for a single parameter}

Let's consider the simplest type of hypothesis to test: a hypothesis about a single parameter. We are interested in to evaluate whether there is an effect of the variable $X_1$ only on the first response. The hypothesis can be written as follows:

\begin{equation}
\label{eq:ex1}
H_0: \beta_{11} = 0 \ vs \ H_1: \beta_{11} \neq 0.
\end{equation}

>>>>

\documentclass{article}
\usepackage{float}
\usepackage{multirow}
\begin{document}
%--------------------------------------------------------------
